---
title: 'Linking movement to spatial covariates through the Langevin diffusion.'
subtitle: 'Tutorial for the Moving2Gather workshop, March 2020'
author: "Pierre Gloaguen"
date: "March 13th, 2020"
output: 
  html_document:
    toc: yes
    toc_float: yes
    theme: journal
    number_sections: yes
bibliography: bibliography.bib
---

```{r setup, include = FALSE}
rm(list = ls())
knitr::opts_chunk$set(echo = TRUE, comment = NA, cache = TRUE)
```


# About this tutorial {-}

This tutorial explains how to fit the Langevin diffusion movement model published in @michelot2019langevin using the current version of the `R` package [@rcore2019language] `Rhabit`.

As this package might evolve, some of the following codes may not work in the future. Feel free to contact me at pierre.gloaguen@agroparistech.fr to get up to dates versions.

# Getting the required packages {-}

This tutorial mostly rely on the `Rhabit` package, that is currently available on [Github](https://github.com). 

## Installation of `Rhabit` {-}

`Rhabit` can be installed directly from there, typing in the `R` console (`devtools` is an `R` package that you might need to install):

```{r install_rhabit, eval = FALSE}
devtools::install_github("papayoun/Rhabit") # Installing (run only once)
```

The installation might require installation (or updating) of other packages.

## Loading required packages {-}

In the following (in addition to the specific `Rhabit` functions), all data cleaning, preparation and visualization will be done using a bunch of packages gathered in the metapackage `tidyverse`.
This regroups a set of convenient packages  for data analysis (see the [R for data science](https://r4ds.had.co.nz/) book).

```{r load_packages, message = FALSE, cache = FALSE}
library(Rhabit) # Package dedicated to the Langevin movement model
library(tidyverse) # Set of useful packages, including ggplot2
library(raster)
```

# The Langevin movement model

In this section, we suppose that we have a set of observed (GPS) trajectories, and we want to link the observed movement with spatial covariates.

## Generic formulation 

The Langevin movement model of @michelot2019langevin belongs to the wider class of continuous time potential-based movement models, mainly proposed by David Brillinger in a series of articles (see @brillinger2010handbook, chapter 26, and the references therein)

The position process $\left\lbrace X_t \right\rbrace_{t \geq 0}$ (taking values in $\mathbb{R}^2$) is supposed to be the stochastic process solution to the following stochastic differential equation:
$$
\text{d} X_t = \frac{\gamma^2}{2}\nabla_x \log \pi_\theta(X_t) \text{d} t + \gamma \text{d} W_t 
$$

where:

- $\pi_\theta(x): \mathbb{R}^2 \mapsto \mathbb{R}$ is a stricly positive function, depending on some unknown parameter $\theta$. This function can be interpretated as the *utilization distribution* (see below), or the eventual stationnary distribution of the individual's position.
- $\gamma$ is an unknown *speed* parameter, reflecting the fact that individuals having different speeds can have a same utilization distribution.
- $\nabla_x$ is the gradient operator.

## About the function $\pi_\theta(x)$ and the process $\left\lbrace X_t \right\rbrace_{t \geq 0}$.

The stochastic differential equation (SDE) defined above is known as the *Langevin diffusion* [@roberts1996exponential].

As it is an SDE, its solution is a **continuous time Markov process with continuous paths in $\mathbb{R}^2$**. 
Moreover, a remarkable property of the Langevin diffusion is that if $\pi_\theta(x)$ is a a probability density function, then its is the *stationnary distribution* of $\left\lbrace X_t \right\rbrace_{t \geq 0}$.
This means that, when $t$ increases, for any area $\mathcal{A}$ of $\mathbb{R}^2$:
$$\mathbb{P}(X_t \in \mathcal{A}) \underset{t \rightarrow \infty}{=} \int_{\mathcal{A}}\pi_\theta(z) \text{d}z\\$$

In movement ecology, this translates to: $\pi_\theta(\cdot)$ is the **utilization distribution** [@worton1989kernel] of the considered individual.

## Linking $\pi_{\theta}$ to spatial covariates: resource selection functions

In ecology, a common obective is to link the utilization distribution to spatial covariates.
Namely, suppose that you have a set of $J$ spatial covariates, modelled through smooth (thus, with derivatives) functions from $\mathbb{R}^2$ to $\mathbb{R}$, denoted $c_1(x),\dots, ,c_J(x)$. 
A *resource selection function* is function $f_\theta(x)$ linking $\pi_\theta(x)$ to $c_1(x),\dots, c_J(x)$:

$$\pi_\theta(x) = f_\theta(c_1(x),\dots, c_J(x)).$$

A common parametric form of $f_\theta(x)$, having an easy interpretation is the following:
$$\pi_\theta(x) = \frac{\exp{\sum_{j = 1}^J}\beta_jc_j(x)}{\int_{\mathbb{R}^2}\exp{\sum_{j = 1}^J}\beta_jc_j(z)\text{d}z} \propto \exp{\sum_{j = 1}^J}\beta_jc_j(x).$$
In this context $\theta = \lbrace \beta_1,\dots, \beta_J \rbrace$, they are coefficients related to the importance of each covariates in the probability of presence of the individual. The denominator is  simply a constant term ensuring that $\pi_\theta$ integrates to 1, and thus is a p.d.f.

**This is the specific resource selection function that is encoded in the `Rhabit` package**.

## Formulation of the Langevin model with exponential RSF

Finally, if we plug the above expression of $\pi_\theta(x)$ in the Langevin movement model, this results in the following movement model:

$$
\text{d} X_t = \frac{\gamma^2}{2}\left(\beta_1 \nabla_x \log c_1(X_t) + \dots +  \beta_J \nabla_x \log c_J(X_t) \right) \text{d} t + \gamma \text{d} W_t 
$$

This continuous time formulated model therefore links the displacement to environmental covariates, moreover, its formulation ensures, under mild condition, that the resulting stochastic process has a stationnary distribution (which can be interpreted as the utilization distribution).

## Estimation of unknown parameters

Unknown parameters in this model are $\theta = \lbrace \beta_1,\dots, \beta_J \rbrace$ and $\gamma$. 
Using standard approximations for stochastic differential equations (namely, the Euler approximation, see @michelot2019langevin for details), these quantities can be estimated by maximimizing an explicit function.
In the case of the RSF shown above, this maximisation is equivalent to finding the estimates in a linear model, it is thus explicit and straightforward to compute.
Moreover, (approximate) confidence intervals can be obtained as in the linear model.

# Fitting the Langevin movement model with the `Rhabit` package

## A simulated case

We first focus on a simulation case, this will enables us to get familiar with the data format, `Rhabit` functions and outputs, and to see some model checking.

### Trajectory data set {-}

Let's first import the data set and print it's first lines:
```{r trajectory_data, echo = -2}
trajectory_data <- read.table(file = "tutorial_simulated_data.txt",
                              sep = "\t", header = TRUE,
                              colClasses = c(ID = "factor")) # ID column is a factor
knitr::kable(head(trajectory_data))
```


This trajectory data set must be a data.frame having 3 mandatory columns:

- `t`: The recording time of the position;
- `x`: The $x$-component of the position;
- `y`: The $y$-component of the position;

In case of multiple individuals/trajectories per individual, an `id` column can be given.

This data set consists in 20 different trajectories, that we will consider as independant realizations of a same movement model.

```{r summary_trajectory_data, comment = NA}
str(trajectory_data)
```


The resulting data set can be represented using usual `ggplot` code:

```{r plotting_simulated_data}
ggplot(trajectory_data) + # Data set which will be plotted
  aes(x = x, y = y) + # Axis
  geom_path(aes(color = ID)) + # Plotting trajectories, One color per trajectory
  # We add starting points
  geom_point(data = dplyr::filter(trajectory_data, t == 0), col = "blue") + 
  theme(legend.position = "none") + # No need for a legend
  labs(x = "X-axis", y = "Y-axis", title = "Trajectory data set")
```

### Covariates

Together with this trajectory data set, two environmental covariates were measured on the area:

```{r environment_covariates}
environment_covariates <- get(load("environment_covariates.RData"))
```

The `environment_covariates` object is a list (here, of length 2). 

```{r properties_env_cov}
is(environment_covariates) # A list
(J <- length(environment_covariates)) # J is the number of covariates
```

Each element of the list is a "raster like" object.
Basically, this means that each covariates consists in a grid (regular or not) on which covariates values are given. In `Rhabit` this means that each covariates is a list having three elements

- `x`: a strictly increasing vector of values (the $x$ values of the grid), of size, say, $n_x$;
- `y`: a strictly increasing vector of values (the $y$ values of the grid), of size, say, $n_y$;
- `z`: A $n_x \times n_y$ matrix of covariates values.

In our example, $n_x = n_y = 201$, and the grid is regularly spaced between -10 and 10.

```{r str_environment_covariates}
str(environment_covariates)
```

We can visualize the covariates using the `plotCovariates` function (in `ggplot`, this requires to transform covariates in a data.frame).

```{r plot_covariates}
plotCovariates(environment_covariates)
```

Our question is whether of not the movement is driven by these covariates? If so, how much? 
From this displacement, can we infer the utilization distribution of this individual?

As a first idea, we can superpose the trajectory data to the previous plot:

```{r plot_covariates_plus_traj, echo = FALSE}
# Same plot, adding trajectory
plotCovariates(environment_covariates, trajectory_data = trajectory_data)
```

To answer these questions, we will use the Langevin movement model.

### Side remark about the plots {-}

The plot functions are aimed at "accelerating" the visualization. Of course,
you might want to transform the covariates data in a `data.frame` to plot in your own favorite way.
This can be easily done using the `rasterToDataFrame` function (whose name might be poorly chosen so far).

```{r get_covariates, eval = FALSE}
# Gathering all covariates in data.frame
covariates_df <- map_dfr(environment_covariates, # To each element of the list, apply
                         Rhabit::rasterToDataFrame, # This function
                         .id = "Covariate")
```

## Computing covariates gradient at sampled locations

As seen in Section 1, the model stipulated that the displacent depends on the gradient of the covariates, and not on the covariates themselves.
Therefore, from our covariates, we must obtain a value of their gradients at each sampled location.

This can be done by any of your favorite method. 
As a possible solution, the `Rhabit` package proposes the function `bilinearGradArray`.
This functions takes the matrix of sampled locations (of size $n \times 2$) and the covariates rasters and returns a $n \times 2 \times J$ array giving all the needed gradient vectors.
The method used is basic bilinear interpolation on a grid (see @michelot2019langevin).

```{r gradient_array}
locations <- trajectory_data %>% # From trajectory_data, extract 
  dplyr::select(x, y) %>% # the  sampled locations  (the x and y columns)
  as.matrix() # and convert it to a matrix 
gradient_array <- Rhabit::bilinearGradArray(locs = locations,
                                            cov_list = environment_covariates)
```

## Fitting the model

Now, we have all the ingredients to fit the Langevin movement model to the data.
This is done with the function `langevinUD`.
The fitting is done using the explicit least squares estimator of the linear model. Thus, for reasonnable amount of data, this should be really quick.

```{r langevin_fitting}
langevin_fit <- Rhabit::langevinUD(locs = locations, # Matrix of locations 
                                   times = trajectory_data$t, # Vector of times
                                   ID = as.character(trajectory_data$ID), # Vector of IDs
                                   grad_array = gradient_array) # Gradient at locations
```

The results are structured into a list having different elements:

```{r langevin_fit_elements}
str(langevin_fit)
```

## Inspecting estimates

### Estimates and confidence intervals {-}

The first interest that we have is in the values of the $\beta$ coefficients.
The values are obtained in the `betaHat` elements.

```{r betaHat_first_model}
langevin_fit$betaHat
```

One can see that the first two coefficients have higher values than the third one. 
As usual, having confidence intervals would allow us to decide whether these values are non zero. 
Asymptotic approximate 95% (the confidence region can be set differently in the `langevinUD` function) confidence intervals are given in the `CI` element.

```{r CI_first_model}
langevin_fit$CI
```

One can therefore see that none of the 95% confidence intervals covers the 0 value.

### AIC criterion

The inference procedure rely on a pseudo log-likelihood (obtained through the Euler discretization). Therefore, a pseudo AIC can be computed to compare different models. 

```{r AIC_first_model}
langevin_fit$AIC
```

## Inspecting residuals

As the model boils down to a linear model (under our approximation assumption).
We can inspect the predicted values and the residuals.

Note that the residuals are the residuals of the normalized increments (the $\mathbf{Y}$ vector in @michelot2019langevin), whereas, the predicted are the predicted positions given the last observations and the model estimates. 
Both these elements are `tibble` having the same number of rows than the original data.

```{r residuals}
langevin_fit$residuals
```

```{r predicted}
# Note that the first value of each trajectory cannot be predicted
langevin_fit$predicted
```

One can therefore see whether the expected displacement matches the actual one.

On the following graph, we show the first 30 points of the first trajectory (the purple dot marks the first position).

```{r plot_predicted}
langevin_fit$predicted %>% 
  rename(x_pred = x, y_pred = y) %>%
  select(x_pred, y_pred) %>% 
  bind_cols(trajectory_data) %>% 
  filter(ID == "1") %>% 
  slice(1:30) %>% 
  ggplot(aes(x = x, y = y)) + 
  geom_point() +
  geom_point(data = slice(trajectory_data, 1), # Mark the first point
             col = "purple", size = 5)+ 
  geom_path(aes(group = ID), color = "blue", linetype = 2) + # Plot the path
  geom_segment(aes(xend = lead(x_pred), # Plot the expected displacement
                   yend = lead(y_pred)), # With arrows
               color = "red", 
               arrow = arrow(length = unit(0.01, "npc")))
```

One can see that the expected displacement is way underestimated.

## Estimated utilization distribution

Using the `plotUD` function, we can visualize the estimated UD.

```{r plot_ud}
Rhabit::plotUD(environment_covariates, langevin_fit$betaHat) +
  # geom_path(data = trajectory_data, aes(group = ID)) + # If you want to add trajectories
  labs(x = "X-axis", y = "Y-axis",
       title = "Estimated UD")
```

# Exercise

Add to the current list of covariates the `distance_covariate` that can be loaded:
```{r get_all_covariates}
distance_covariate <- get(load("distance_covariate.RData"))
all_covariates <- c(environment_covariates,
                    distance_covariate)
```

This covariate simply computes the squared distance to the center of the map (the point $(0, 0)$).

Following the previous codes: 

- Fit the Langevin Movement model using this new set of covariates.
- Check again the estimates of the $\beta$ coefficients.
- Choose the best model (using the AIC criterion).
- Check again the predicted displacement to see whether predictions are improved.

# Application to Stellar sea lion

In this section, we'll fit the Langevin model to actual stellar sea lion data.

The data was published in @wilson2018estimating and consists in the trajectories of 3 stellar sea lions around the Aleutians islands in Alaska (USA).
The data was preprocessed to remove GPS error (see @michelot2019langevin for details).

## Loading and transforming trajectory data

We load the data, the time column is in the classical "POSIXct" format.

```{r seal_tracks}
seal_tracks <- read.table(file = "seal_data.csv", header = T, sep = ",",
                          colClasses = c(ID = "factor", # Specifying column classes
                                         time = "POSIXct")) %>% 
  mutate(x = x / 1000,
         y = y / 1000)
```

The data are projected in a certain projection and can be plotted:

```{r plotting_ssl_data}
ggplot(seal_tracks) +
  aes(x = x, y = y, col = ID) +
  geom_path() +
  labs(x = "Easting (meters)", 
       y = "Northing (meters)")
```

`Rhabit` do not handle `POSIXct` but an increasing vector of numeric times.

```{r head_seal_tracks}
seal_tracks_Rhabit <- seal_tracks %>% # 
  group_by(ID) %>% # Applying same treatment to each group
  # Changing the time column, in an increasing numeric vector (here in hours)
  mutate(time = c(0, cumsum(diff(as.numeric(time))))) %>% 
  ungroup()
```

## Loading and transforming covariates

Originally, the covariates are stocked in a raster directory (having two files, a `.grd` and a `.gri` file). 
We want to load it (using the `raster` package) and then transform it to a list of lists, as seen above.
This transformation is done using the function `rasterToRhabit`.

```{r loading_ssl_covariates}
habitat_covariates <- raster::brick("habitat_seal_data/aleut_habitat.grd", 
                                    values = TRUE) %>%
  # Raster brick, in some projection
  raster::as.list() %>% # Turns it into a list
  purrr::map(Rhabit::rasterToRhabit) %>%  # To each list element, we apply the rasterToRhabit function
  purrr::map(function(covariable){
    covariable$x <- covariable$x / 1000
    covariable$y <- covariable$y / 1000
    return(covariable)
  })
```

Then you can plot it (quite long...)

```{r plot_ssl_covariates}
Rhabit::plotCovariates(habitat_covariates)
```

## Fitting the model on data

This is as simple as you expected it to be!

```{r fitting_langevin_sll}
# First, get locations as a matrix
locations_ssl <- seal_tracks_Rhabit %>% # From trajectory_data, extract 
  dplyr::select(x, y) %>% # the  sampled locations  (the x and y columns)
  as.matrix() # and convert it to a matrix 
# Then, get gradients at each gradient array
gradient_array_ssl <- Rhabit::bilinearGradArray(locs = locations_ssl,
                                            cov_list = habitat_covariates)
# Then, fit it!
langevin_fit_ssl <- langevinUD(locs = locations_ssl,
                               times = seal_tracks_Rhabit$time,
                               ID = seal_tracks_Rhabit$ID, 
                               grad_array = gradient_array_ssl)
```

## Estimates

We can observe $\hat{\beta}$:

```{r estimates_ssl}
langevin_fit_ssl$betaHat
```

Values are small, we must check the confidence intervals:

```{r CI_ssl}
langevin_fit_ssl$CI
```

It might be of interest to see the correlation matrix of estimates:

```{r cor_estimates}
# The variance-covariance matrix is transformed to a correlation matrix
cov2cor(langevin_fit_ssl$betaHatVariance)
```

Of course, you can observed the estimated utilization distribution:

```{r plot_UD}
plotUD(covariates = habitat_covariates, beta = langevin_fit_ssl$betaHat)
```

## Model fitting

We can check the predicted displacements and compare it to the actual displacements to assess goodness of fit:

```{r checking_predicted_ssl}
langevin_fit_ssl$predicted %>% 
  rename(x_pred = x, y_pred = y) %>%
  select(x_pred, y_pred) %>% 
  bind_cols(seal_tracks) %>% 
  filter(ID == "14809") %>% 
  ggplot(aes(x = x, y = y)) + 
  geom_point() +
  geom_path(aes(group = ID), color = "blue", linetype = 2) + # Plot the path
  geom_segment(aes(xend = lead(x_pred), # Plot the expected displacement
                   yend = lead(y_pred)), # With arrows
               color = "red", 
               arrow = arrow(length = unit(0.01, "npc"))) +
  coord_cartesian(xlim = c(-1780, -1760), ylim = c(560, 590))
```


# References